Modeling memory and taxis to analyze migrations

### 2021-04-10 text/ms.Rmd
Changed the model Wrote some text Copied CWD shiny app

---
title: "Modeling memory and taxis to analyze migrations"
author: "E. Gurarie, S. Potluri, B. Fagan, others"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: yes
  html_document:
    toc: yes
    toc_float: yes
theme: cerulean
editor_options:
  chunk_output_type: console
bibliography: ../library/memorymigration.bib
csl: ecology.csl
---

@@

We intend to study how a blending of a *tactical* (i.e. direct response to resource availability or perception) and a *strategic* (i.e. memory-driven and forward-thinking) behavior can help a forager navigate a dynamic and heterogeneous resource environment, with a particular emphasis on the aquisition and adaptabilty of **migratory behavior**.  
Our model is an advection diffusion with two advective terms corresponding to a direct response to the environment and to memory.  Importantly, the environment is *approximately periodic*, i.e. one in which the distribution of the resource at a given time is approximately equal to that resource distribution at a fixed previous time period, i.e. $h(t) \approx h(t - \tau)$. This models annual variations, and the period corresponds to one year.  The *tactical* process is straightforward taxis up a resource gradient, while the *collective memory* corresponds to a tendency of the organisms to repeat the behavior of the population in previous years.  Thus:

$${\frac{\partial u(x,t)}{\partial t}} = -\varepsilon {\frac{\partial^2 u(x,t)}{\partial x^2}} + \alpha \frac{\partial}{\partial x}\left(u \frac{\partial h(x,t)}{\partial x}\right) + \sum_{i = 0}^n \beta_i \frac{\partial u(x, t - i\tau)}{\partial x}$$
where $u$ represents a population distributed in time and space, $\varepsilon$ is a rate of randomness (diffusion), $\alpha$ is the strength of advection in the direction of the resource $h(x,t)$ gradient), and $\beta_i$ is the strength of advection up the gradient of the *population* distribution in an $i$'th earlier seasonal cycle (year), up to $n \in N$.

Note that $\beta_0$ is similar to an index of "cohesion", essentially encouraging the population to stay close together.  Higher terms of $\beta_i$ should be simple, e.g., if the collective memory reaches only one year back then $n = 1$ and $\beta_i$ is a single constant. Alternatively, a form of memory might discounts older "traditions" via $\beta_i = \beta \exp(-i/\gamma)$, where $\gamma$ represents a generational scale of memory.

>>>>

---
title: "Modeling memory and taxis to analyze migrations"
author: "E. Gurarie, S. Potluri, B. Fagan, others"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: yes
theme: cerulean
editor_options:
  chunk_output_type: console
bibliography: ../library/memorymigration.bib
csl: ecology.csl
---

@@

We study how a blending of a *tactical* (i.e. direct response to resource availability or perception) and a *strategic* (i.e. memory-driven and forward-thinking) behavior can help a forager navigate a dynamic and heterogeneous resource environment, with a particular emphasis on the aquisition and adaptabilty of **migratory behavior** in a dynamic, seasonal resource environment, i.e. one in which the distribution of the resource at a given time is approximately equal to that resource distribution at a fixed previous time period, i.e. $h(t) \approx h(t - \tau)$.

Our model is an advection diffusion with two advective terms corresponding to a direct response to the environment and to memory. The *tactical* process is straightforward taxis up a resource gradient, while the *collective memory* corresponds to a tendency of the organisms to repeat the behavior of the population in previous years.  Thus:

$${\frac{\partial u(x,t)}{\partial t}} = -\varepsilon {\frac{\partial^2 u(x,t)}{\partial x^2}} + 
\alpha \frac{\partial}{\partial x}\left(u(x,t) \frac{\partial h(x,t)}{\partial x}\right) + 
\beta  \frac{\partial}{\partial x}\left(u(x,t) \frac{\partial m(x,t)}{\partial x}\right) \,\,\,\,(1)$$

where the variables $u(x,t)$ represents a population distributed in time and space, $h(x,t)$ is the distribution of the resource, $m(x,t,y)$ is the memory component in space and time at (discrete) year $y$, $\varepsilon$ is a rate of randomness (diffusion), $\alpha$ is the strength of advection in the direction of the resource $h(x,t)$ gradient), and $\beta$ is the strength of advection up the gradient memory.  The memory component we model as a mixture of a reference (long-term) memory, modified by more recent experiences as follows:
$$m(x,t) = \kappa^y \, u(x, t - y\tau) + \left(1-\kappa^y\right)\,u(x, t-\tau)\,\,\,\,(2)$$
Here, $y \in \{0,1,2,...\}$ represents the count of a given year, i.e. $y = \lfloor t/\tau \rfloor$. Thus, $u(x, t - y\tau)$ is the initial (year 0) "learned" collective behavior, $u(x, t-\tau)$ is the preceding year's collective behavior and $\kappa \in (0,1)$ is the relative strenfth of the long-term (reference) memory.   When $\kappa = 1$, all of the memory is reference memory, and the system essentially responds to it's initial learned behavior as if it were a resource. If $\kappa = 0$, the memory is short-term, and the process responds only to the behavior from the previous year.  With $0 < \kappa < 1$, the reference memory fades with time.
